---
title: "Assignment 00"
author: "Krista Pipho, kp275"
date: "January 25, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#First Assignment
##Illustrating R Markdown Features


###Formatting

Starting with an un-ordered list

*
We can make **Bold** text
    - Also strike <del>WOW
*
We can make *Italic* text
    - Also superscript SO^FANCY^
*
We can make `verbatim` text
    - Also subscript SO~SHINY~


Then an ordered list

1. Here is a [Hyperlink](https://bio723-class.github.io/Bio723-assignments/00-assignment.html)
2. Here is a Footnote^[Nifty]
    - Additional Footnote^[Cool]
  

###LaTeX Equations

Mean

$\bar x = \frac{1}n \sum_{i=1}^{n} x_i$

Variance

$s_x^2 = \frac1{n-1} \sum_{i=1}^{n} (x_i - \bar x)^2$

Standard deviation

$s_x = \sqrt{s_x^2}$

Covariance 

$s_{x,y} = \frac{1}{n-1} \sum_{i=1}^{n}(x_i - \bar x)(y_i - \bar y)$

Correlation

$r_{x,y} = \frac{s_{x,y}}{s_x s_y}$

###Basic Functions


```{r}

#Define Vector 
x = c(58:72)


#Calculate mean, variance, and standard deviation

mean(x)
var(x)
sd(x)

```

```{r}

#Define second vector
y = c(115,117,120,123,126,129,132,135,139,142,146,150,154,159,164)

#Calculate covariance and correlation
cov(x,y)
cor(x,y)
  
```



