---
title: "Bio 723: Pipho_K_Assignment03"
author: "Krista Pipho, kp275"
date: "February 17, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Homework 3

###Problem 1. 

Write a function, vec.dot(x,y), that calculates the dot product of two vectors, x and y, of arbitrary number of elements.

```{r}

a <- 1:15
b <- 10:24
c <- 1:24

vec.dot <- function(x,y){
dot <- 0
if(length(x) == length(y)){
for(i in 1:length(x)){
  dot <- dot + x[i]*y[i]}
return(dot)}else{return("Error, lengths mismatched")}
}  

# Demonstrating output of the function I wrote
vec.dot(a,b)
# Demonstrating output of my function when the vectors have mismatched lengths
vec.dot(a,c)
# Showing that my output is the same as that for built in functions
sum(a*b)

```



###Problem 2
Write a function, vec.length(x), that calculates the Euclidean length of a vector.

```{r}

vec.length <- function(x){
  count <- 0 
  for(i in 1:length(x)){
    count <- count + x[i]*x[i]}
  count = sqrt(count)
  return(count)
}

# Demonstrating my function for vector length
vec.length(a)

# A deffinition using built in functions to calculate vector length
test.length <- function(x){sqrt(sum(x^2))}

# Result of built in functions for comparasin 
test.length(a)

  
```

###Problem 3
Write a function vec.cosine(x,y), that calculates the cosine of the angle between two vectors using your vec.dot() and vec.length() functions.

```{r}
library(lsa)

vec.cosine <- function(x,y){vec.dot(x,y)/(vec.length(x)*vec.length(y))}

# Demonstrating my function for calculating cosine of the angle between two vectors
vec.cosine(a,b)

# Result of built in function for comparasin 
cosine(a,b)

```

###Problem 4

Write a function vec.angle(x, y, deg=FALSE) that by defaults calculates the angle, in radians, between two vectors. When the argument deg=TRUE your function should return the angle in degrees.

```{r}

vec.angle <- function(x,y,deg = FALSE){
  x.cos <- vec.cosine(x,y)
  if(deg == FALSE){return(acos(x.cos))}
  if(deg == TRUE){return((acos(x.cos)*180)/pi)}
}

vec.angle(a,b, TRUE)

```

###Problem 5

Write a function, vec.distance(x,y) that calculates the distance between two vectors (i.e. distance between their end points)

```{r}

vec.distance <- function(x,y){
  len.x <- vec.length(x)
  len.y <- vec.length(y)
  cos.x.y <- vec.angle(x,y)
  c <- sqrt(len.x^2 + len.y^2 - 2*(len.x)*(len.y)*(cos.x.y))
  return(c)
}

vec.distance(a,b)

```

###Problem 6
Write a function vec.mean(x) that calculates the mean of x, using vector algebraic operations.

```{r}

vec.mean <- function(x){
  sum = 0
  for(i in 1:length(x)){
    sum = sum + x[i]
  }
  mean <- sum/length(x)
  return(mean)
}

vec.mean(a)

```


###Problem 7

Write a function vec.cov(x,y) that calculates the covariance of x and y, using vector algebraic operations. Remember to mean center! Compare the results with built-in cov() function to make sure your function works properly.

```{r}

vec.cov <- function(x,y){
  if(length(x) == length(y)){
  mean.x <- vec.mean(x)
  mean.y <- vec.mean(y)
  sum = 0
  for(i in 1:length(x)){
   sum = sum +(x[i]-mean.x)*(y[i]-mean.y)
  }
  return(sum/(length(x)-1))}
  else{return("Error, lengths mismatched")}
}

# Showing the output of my function for calculating covariance
vec.cov(a,b)

# Result of built in function for comparasin 
cov(a,b)

```


###Problem 8

Write a function vec.var(x) that calculates the variance of x, using vector algebraic operations. Remember to mean center! Compare the results with built-in var() function to make sure your function works properly.

```{r}

vec.var <- function(x){
  mean.x <- vec.mean(x)
  sum = 0
  for(i in 1:length(x)){
    sum <- sum + (x[i]-mean.x)^2
  }
  return(sum/(length(x)-1))
}

# Showing the output of my function for calculating variance 
vec.var(a)

# Result of built in function for comparasin 
var(a)

```


###Problem 9
Write a function vec.cor(x,y) that calculates the correlation of tow vectors, x and y, using vector algebraic operations. Remember to mean center! Compare the results with built-in cor() function to make sure your function works properly.

```{r}

vec.cor <- function(x,y){

  cov.x.y <- vec.cov(x,y)
  sd.x <- sd(x)
  sd.y <- sd(y)
  
  cor <- cov.x.y/(sd.x*sd.y)
  return(cor)
}

# Showing the output of my function for calculating corelation  
vec.cor(a,b)

# Result of built in functions for comparasin 
cor(a,b)

```








