---
title: "Bio 723: Pipho_K_Assignment04"
author: "Krista Pipho, kp275"
date: "February 24, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Homework 4

###Problem 1

Write a function, mean.vector(X), that takes a matrix of real values (doubles) and returns the column means as a vector. Do not use built-in vector functions such as sum(), mean(), colSums(), etc.

```{r}

r.m <- matrix(rnorm(15), nrow=5)

mean.vector <- function(x){
  means <- vector()
  
  for(i in 1:dim(x)[2]){
    sum = 0
    for(j in 1:dim(x)[1]){
      sum <- sum + x[j,i]}
    means <- c(means, sum/j)}
  return(means)}

mean.vector(r.m)
```


###Problem 2

Write a function, deviation.matrix(X), that takes a matrix of real values and returns a matrix the represents each values deviation from it's respective column mean. Do not use built-in functions such as scale().

```{r}

deviation.matrix <- function(x){
  m <- mean.vector(x)
  all.deviations <- vector()
  for(i in 1:dim(x)[2]){
    deviations <- vector()
    for(j in 1:dim(x)[1]){
      deviations <- rbind(deviations,(x[j,i]-m[i]))
    }
    all.deviations <- cbind(all.deviations,deviations)
  }
  return(as.matrix(all.deviations))
}

deviation.matrix(r.m)

```

###Problem 3

Write a function covariance.matrix(X), that takes as input a matrix of real values (variables in columns) and returns a covariance matrix. Compare the results with built-in cov() function to make sure your function works properly.

```{r}

covariance.matrix <- function(x){
  x <- deviation.matrix(x)
  all.covs <- vector()
  for(i in 1:dim(x)[2]){
    covs <-  vector()
    for(h in 1:dim(x)[2]){
      sum  = 0
      for(j in 1:dim(x)[1]){

        sum <- sum + x[j,i] * x[j,h]
      }
      covs <- rbind(covs,sum/(j-1))
    }
    all.covs <- cbind(all.covs,covs)
  }
  return(all.covs)
  }

#Show results from my covariance matrix function
covariance.matrix(r.m)
#Compare with results from built in covariance matrix function
cov(r.m)

```

###Problem 4

Write a function correlation.matrix(X), that takes as input a matrix of real values (variables in columns) and returns a correlation matrix. Compare the results with built-in cor() function to make sure your function works properly.

```{r}

correlation.matrix <- function(x){
  dev.m <- deviation.matrix(x)
  means.m <- mean.vector(x)
  cov.m <- covariance.matrix(x)
  all.sd <- vector()
  for(i in 1:dim(dev.m)[2]){
    sum = 0
    for(j in 1:dim(dev.m)[1]){
      sum <- sum + (dev.m[j,i])^2
    }
    all.sd <- c(all.sd,sqrt((sum/(j-1))))
  }
  all.cor <- vector()
  for(i in 1:dim(x)[2]){
    cors <-  vector()
    for(j in 1:dim(x)[2]){
      
     cors <- rbind(cors,cov.m[j,i]/(all.sd[j]*all.sd[i]))
    }
    all.cor <- cbind(all.cor,cors)
  }
  return(all.cor)
}

#Showing output of my correlation matrix function
correlation.matrix(r.m)
#Comparing to the output of the built in correlation function
cor(r.m)

```
