---
title: "Pipho_K_Assignment05"
author: "Krista Pipho, kp275"
date: "March 15, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##Bivariate Regression 

Load in the lion nose color dataset

```{r, message=FALSE, warning = FALSE}

library(tidyverse)
library(ggExtra)

url <- "https://github.com/bio304-class/bio304-course-notes/raw/master/datasets/ABD-lion-noses.csv"
nose_color <- read_csv(url)
glimpse(nose_color)

```

###Problem 1 a:

Generate a bivariate plot (with marginal histograms) showing the relationship between nose pigmentation (the predictor variable) and age (the dependent variable) 

```{r}

bi_var_marginal_histograms <- ggplot(nose_color, aes(x = proportionBlack, y= ageInYears)) + geom_point() + labs(x = "Proportion Black", y = "Age", 
       title = "Bivariate Plot of Nose Pigmentation vs. Age")
ggMarginal(bi_var_marginal_histograms, type = "histogram", bins = 11)

bi_var_marginal_histograms

```


###Problem 1 b:

Using vector operations calculate the regression coefficients for the regression of age on nose pigementation.

```{r}

# Calculating mean centered vectors
centered_proportionBlack <- nose_color$proportionBlack - mean(nose_color$proportionBlack)
centered_age <- nose_color$ageInYears - mean(nose_color$ageInYears)

# Estimate coefficients
b.est <- (centered_proportionBlack %*% centered_age) / (centered_proportionBlack %*% centered_proportionBlack)
a.est <- mean(nose_color$ageInYears) - b.est * mean(nose_color$proportionBlack)

# 'drop' used to convert to scalar values
regression <-  list(intercept = drop(a.est), slope = drop(b.est))
regression

```

###Problem 1 c:

Modify your bivariate plot for problem 1a to illustrate the estimated regression.


```{r}

bi_var_marginal_histograms_abline <-  bi_var_marginal_histograms + geom_abline(slope = regression$slope, intercept = regression$intercept, color = "red")
ggMarginal(bi_var_marginal_histograms_abline, type = "histogram", bins = 11)
bi_var_marginal_histograms_abline


```

###Problem 1 d:

####Based on the regression of age on proportion of black pigmentation, if you observed a male lion with no black pigmentation on its nose, how old would you predict it to be?

The y intercept (0,y) is .8790062, thus when the proportion black on the nose (x) is zero we would predict that the animals age is 12*0.8790062 (~10.8) months old. 


###Problem 1 e:

####What is the predicted increase in age for a 10% increase in black pigmentation? 

The slope (x/y) is 10.64712, thus for each increment of 10% increase in nose pigmentatin we would predict an age increase of 10.64712*0.1 (1.06) years. 


###Problem 1 f:

What is the coefficient of determination for your fit model? 

```{r}

# Calculating goodness of fit, or coefficient of determination
age_hat <- regression$intercept + regression$slope * nose_color$proportionBlack

age_hat.ctr <-age_hat - mean(age_hat)

coeff.determination <-  drop(age_hat.ctr %*% age_hat.ctr) / drop(centered_age %*% centered_age)
coeff.determination

```

###Problem 1 g:

Generate a residual plot to examine the residuals from the estimated regression. Are the residuals approximately normal? Are the evenly distributed around zero? Are the any signs of trends in the residuals?


```{r}
# Calculating the residuals

e <- nose_color$ageInYears - age_hat

regression.df <- tibble(X = nose_color$proportionBlack, Y = nose_color$ageInYears, fitted = age_hat, residuals = e)


# Making a residual plot
ggplot(regression.df, aes(X,residuals)) + geom_point() + geom_hline(yintercept = 0, color = 'red', linetype = 'dashed') + labs(x = "Proportion Black", y = "Residuals", title = "Residual Plot for Nose Pigmentation vs. Age")


```

The residuals do not appear normally distributed, rather they look skewed to the left. There seems to be an appoximately even distribution around zero. I perceive a trend towrads increasing absolute value of the residuals from points on the right. 


###Problem 2:

Write your own function bivariate.regression(X, Y) that takes as input two vectors, X and Y, and which calculates the regression of Y on X using vector operations. Your function should return a list object that includes two elements

> a list that includes the slope and intercept of the fitted model

> a data frame that includes the X and Y values, the fitted (predicted) values of Y, and the residual values.

```{r}

bivariate.regression <- function(x, y) {
  
# Calculating mean centered vectors
centered_x <- x - mean(x)
centered_y <- y - mean(y)

# Estimate regression coefficients
b.est <- (centered_x %*% centered_y) / (centered_x %*% centered_x)
a.est <- mean(y) - b.est * mean(x)

###Question 2a
# 'drop' used to convert to scalar values, makes a list of fitted model values
regression <-  list(intercept = drop(a.est), slope = drop(b.est))

# Calculating goodness of fit, or coefficient of determination
y_hat <- regression$intercept + regression$slope * x

# Calculating the residuals
e <- nose_color$ageInYears - y_hat

###Question 2b
# Makes a dataframe of the two original vectors, the fitted y (predicted) y values, and the residuals 
regression.df <- tibble(X = x, Y = y, fitted = y_hat, residuals = e)

# Returns the list and table 
return(list(regression = regression, regression.df = regression.df))
}

reg.model <- bivariate.regression(nose_color$proportionBlack, nose_color$ageInYears)
reg.model

```

###Problem 3:

Write a function regression.plots(reg.model) that takes as input the output object for your function in problem 1, and which returns one nicely formatted plot with two subfigures:

>a figure illustrating the bivariate distribution of X and Y and a line representing the regression model. Include in the title of this subplot the values of coefficients of the regression model

>a figure illustrating the residuals from the regression model as a function of X (i.e. a residual plot)

```{r fig1, fig.height = 3, fig.width = 10}
library(cowplot)

regression.plots <- function(reg.model){
regression <- reg.model$regression
regression.df <- reg.model$regression.df

bi_var_marginal_histograms <- ggplot(regression.df, aes(x = X, y= Y)) + geom_point() + labs(x = "Proportion Black", y = "Age", title =  "Bivariate Plot of Nose Pigmentation vs. Age", subtitle = paste("Regression line: y =", regression$slope, "x +",regression$intercept))
bi_var_marginal_histograms_abline <-  bi_var_marginal_histograms + geom_abline(slope = regression$slope, intercept = regression$intercept, color = "red") 
ggMarginal(bi_var_marginal_histograms_abline, type = "histogram", bins = 11)
bi_var_marginal_histograms_abline

# Making a residual plot
residual_plot <- ggplot(regression.df, aes(X,residuals)) + geom_point() + geom_hline(yintercept = 0, color = 'red', linetype = 'dashed') + labs(x = "Proportion Black", y = "Residuals", title = "Residual Plot for Nose Pigmentation vs. Age", subtitle = "")

composite_plot_1 <- ggdraw() + 
  draw_plot(bi_var_marginal_histograms_abline, 0, 0, width = .5) + 
  draw_plot(residual_plot, 0.5, 0, width = .5) +
  draw_plot_label(c("A.", "B. "), c(0, 0.5), c(1, 1), size = 15)

composite_plot_1
}

##Parts a and b labled on figure as 'A' and 'B' respectively
regression.plots(reg.model)

```

