---
title: "Assignment 7"
author: "Krista Pipho, kp275"
date: "March 31, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##PCA and CVA

Load in libraries

```{r message=FALSE, warning = FALSE}

library(tidyverse)
library(cowplot)
library(broom)
library(magrittr)
library(MASS)
library(ggforce)

```


###Problem 1:

Create a data frame from the wine data, and assign meaningful column names

```{r}

url <- "https://raw.githubusercontent.com/Bio723-class/example-datasets/master/wine.data"
wine <- read.csv(url, header = FALSE)

# Name the wine dataset columns
colnames(wine)<- c("Cultivar", "Alcohol", "Malic Acid", "Ash", "Alcalinity of ash", "Magnesium", "Total phenols", "Flavanoids", "Nonflavanoid phenols", "Proanthocyanins", "Color intensity", "Hue", "OD280/OD315 of diluted wines", "Proline")

glimpse(wine)

```


###Problem 2:

Generate a derived data frame with the centered and scaled chemical attributes.

```{r}

measures.only <- wine %>% dplyr::select(-Cultivar)
cultivar.only <- wine %>% dplyr::select(Cultivar)
cultivar.only <- lapply(cultivar.only, as.factor)

#Center and scale te data
measures.only <- as.data.frame(scale(measures.only))

```


###Problem 3:

Using the centered-and-scaled attributes, carry out PCA of the wine data set. Generate a plots of the observations in the space of the first two PC axes. How much of the variation in the data is captured by PCs 1 and 2?

```{r}

# Run PCA analysis
measures.pca <- prcomp(measures.only, center=TRUE, retx=TRUE)

# Extract PCA scores from prcomp output
pca.scores.df <- as.data.frame(measures.pca$x)

# Define a coordinate system for the plot
coord.system <- coord_fixed(ratio=1, xlim=c(-5,5),ylim=c(-5,5))

# Add the cultivar lables back in
pca.scores.lab <- cbind(pca.scores.df,cultivar.only)

# Plot
pc1v2 <-
  pca.scores.lab %>%
  ggplot(aes(x = PC1, y= PC2, color = Cultivar)) +
  geom_point(size = 2, alpha = .8) + 
  coord.system + theme_classic()

pc1v2

# See a summary of how much varriance is explained by each PC
summary(measures.pca)

```
PC1 captures 36% of varriance in the data, while PC2 captures 19%. 


###Problem 4:

Using the centered-and-scaled attributes, carry out CVA of the wine data set using the MASS::lda() function. How much of the between-group variance is captured by the respective canonical variates?

```{r}

# Adds cultivar lables back on
measures.lab <- cbind(measures.only,cultivar.only)

#
measures.lda <- lda(Cultivar ~ ., data = measures.lab)
measures.lda

```

LD1 captures 69% of the between group variance, while LD2 captures 31%. 


###Problem 5:

Generate a basic CVA plot for the wine data set

```{r}

measures.sub <- 
  measures.only %>% as.matrix  # cast to matrix for calculations

# calculate CV scores
CVA.scores <- measures.sub %*% measures.lda$scaling

# create data frame with scores
measures.CV <- data.frame(CVA.scores)
measures.CV$Cultivar <- as.factor(wine$Cultivar)

measures.cva.plot <-
  ggplot(measures.CV, aes(x = LD1, y = LD2)) + 
  geom_point(aes(color=Cultivar, shape=Cultivar), alpha=0.7) + 
  labs(x = "CV1", y = "CV2") +
  coord_fixed(ratio=1) + theme_classic()

measures.cva.plot
```



###Problem 6:

Generate a CVA plot illustrating the 95% confidence regions for the group means. 


```{r}

# Look up the chi squared values for .05 confidance
chi2 = qchisq(0.05,2, lower.tail=FALSE)

# Calculate the chi squared confidance regions
CIregions.mean.and.pop <-
  measures.CV %>%
  group_by(Cultivar) %>%
  summarize(CV1.mean = mean(LD1),
            CV2.mean = mean(LD2),
            mean.radii = sqrt(chi2/n()),
            popn.radii = sqrt(chi2))

# Plot the confidance regions as a for the mean and population
measures.cva.plot2 <-
  measures.cva.plot + 
  geom_circle(data = CIregions.mean.and.pop,
              mapping = aes(x0 = CV1.mean, y0 = CV2.mean, r = mean.radii),
              inherit.aes = FALSE) 

measures.cva.plot2
```


###Problem 7:

Generate another CVA plot illustrating the 95% tolerance regions for the group population distributions. 

```{r}

measures.cva.plot3 <-
  measures.cva.plot +
  geom_circle(data = CIregions.mean.and.pop,
              mapping = aes(x0 = CV1.mean, y0 = CV2.mean, r = popn.radii),
              linetype = "dashed", 
              inherit.aes = FALSE) 

measures.cva.plot3

```


###Problem 8:

Comment on the similarities and differences between the PCA plot and CVA plot of the wine data.

The CVA plot has a very similar shape to the PCA plot, but seperates groups much more clearly. No 'stray' points are present within the 95% confidance interval of a different CVA group, but the edges of PCA groups are quite intermingled. If I needed to descriminate groups using this data I would deffinately choose the CVA over PCA for that analysis. 
